---
title: "Lab Formative Assessment 1"
author: "Sindayen, Lorenzo Danilo V."
subtitle: "DSC1105 Exploratory Data Analysis"
date: "2026-01-25"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
library(ggplot2)
```

## 1. Create a histogram using the diamonds dataset.
## Note: Re-write this plot using the layer() function instead of geom_histogram(), as demonstrated in class.

```{r}
ggplot(diamonds) +
 layer(
     mapping = aes(x = carat),
     stat = "bin",
     geom = "bar",
     position = "stack"
 ) +
ggtitle("1. Histogram of Diamond Carat")
```

## 2. Recall that a histogram is a plot that uses stat_bin rather than stat_identity.

## Note: Modify your histogram from Question 1 so that it uses a different geom, such as geom_line() or geom_point(), while still relying on stat_bin.

```{r}
ggplot(diamonds) +
 layer(
     mapping = aes(x = carat),
     stat = "bin",
     geom = "line",
     position = "identity"
 ) +
ggtitle("2. Histogram using line geometry")
```

## 3. Using the histogram with bars from Question 1, add an aesthetic mapping based on one of the factor variables in the dataset (e.g., cut, color, or clarity).
## Note: Apply this mapping to either the fill or color aesthetic.
```{r}
ggplot(diamonds) +
 layer(
     mapping = aes(x = carat, fill = cut),
     stat = "bin",
     geom = "bar",
     position = "stack"
 ) +
ggtitle("3. Histogram of Diamond Carat by Cut")
```

## 4. What is the default position adjustment used in a histogram?
## Note: Change the position adjustment in the histogram you created in Question 3 to a different option (hint: try dodge), and observe how the plot changes.
```{r}
ggplot(diamonds) +
 layer(
     mapping = aes(x = carat, fill = cut),
     stat = "bin",
     geom = "bar",
     position = "dodge"
 ) +
ggtitle("4. Histogram of Diamond Carat by Cut using (dodge)")
```

