---
title: "Lab Formative Assessment 2"
author: "Sindayen, Lorenzo Danilo V."
subtitle: "DSC1105 Exploratory Data Analysis"
date: "2026-01-29"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(knitr)
library(ggplot2)
```

---

\begin{center}
\textit{GitHub Link:} \\
\url{https://github.com/ChewyGnome/DSC1105}
\end{center}

---

# Dataset
The dataset that will be used in this report is the built-in mpg dataset included in the ggplot2 package. 
```{r}
data(mpg)
```

# Data Inspection

Structure of the dataset:
```{r}
class(mpg)
```

First 6 rows of the dataset:
```{r}
head(mpg)
```

# Data Wrangling and Transformation
```{r}
mpg_wrangle <- mpg %>%
  separate(
    trans,
    into = c("transmission_type", "gears"),
    sep = "\\(",
    remove = FALSE
  ) %>%
  mutate(gears = str_remove(gears, "\\)"))

mpg_wrangle <- mpg_wrangle %>%
  mutate(
    log_hwy = log(hwy),
    sqrt_displ = sqrt(displ)
  )

mpg_wrangle %>% 
  select(trans, transmission_type, gears, log_hwy, sqrt_displ) %>% 
  head()
```

# Visualization

## Histogram of Original Highway Fuel Efficiency
```{r}
ggplot(mpg_wrangle, aes(x = hwy)) +
  geom_histogram(bins = 20) +
  labs(
    title = "Histogram of the original fuel efficiency variable",
    x = "Highway MPG",
    y = "Frequency"
  )
```

## Histogram of the Transformed Highway Fuel Efficiency
```{r}
ggplot(mpg_wrangle, aes(x = log_hwy)) +
  geom_histogram(bins = 20) +
  labs(
    title = "Histogram of the transformed fuel efficiency variable",
    x = "Log(Highway MPG)",
    y = "Frequency"
  )
```

## Q-Q plot of Original Highway MPG
```{r}
ggplot(mpg_wrangle, aes(sample = hwy)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Q-Q Plot of Original Highway MPG")
```

## Q-Q plot of Log-Transformed Highway MPG
```{r}
ggplot(mpg_wrangle, aes(sample = log_hwy)) +
  stat_qq() +
  stat_qq_line() +
  labs(title = "Q-Q Plot of Log-Transformed Highway MPG")
```

# Interpretation
In the data provided, the original highway fuel efficiency exhibits a right skewness in the distribution of data, evident in the both the histogram and the Q-Q plot. However, after the application of the logarithmic transformation, it can be seen that the distribution of the data became a lot more balance. This could also be seen in the Q-Q plot of the transformed variable, where it can be seen that the alignment of the points are a lot more standardized, indicating a decreased deviation of the results.

# Reflection
A log transformation was chosen for highway fuel efficiency due to the fuel economy data showing a skewed distribution of the dataset. The square root transformation of engine displacement helps reduce the influence of large values while preserving interpretability. Concluding this, the transformation of the variable had a massive improvement in normalizing the results. Improvements such as this could be crucial in simplifying the constraints of some certain datasets that rely on consistency and a constant variance.

